Bayesian exponential random graph modeling of KWL Burial network

### 2020-12-18 analysis/paper/paper.Rmd
added a map and references

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

<!-- intro to networks -->
A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison <!-- fig reference to a map of burials at the site-->. Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

## Model specification in a Bayesian framework
```{r ergm-terms-table}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
kableExtra::kable(ergm_terms,
             caption = "The parameters of exponential random graph models for a undirected network to corresponding with archaeological evidence.")
```

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European percicolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  fig.path = "../figures/"
)

image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

<!-- intro to networks -->
A social network is generally visualized as a graph consisting of a set of socially-relevant nodes/actors, connected by edges/ties representing one or more relations, such as friendship, collaborations, information flow, trade ties, or any other forms of connection of interest [@Marin2011; @Wasserman1994]. The ties can be classified into four major types, including similarities, social relations, interactions, and flows [@Borgatti2009]. In archaeology, actors can be people, groups, objects, places, or events, with ties built on similarity, proximity, or co-presence of material culture to create patterns reflecting influence, geographical distance, or affiliations in social groups [@Mills2017; @Brughmans2018; @Peeples2019]. For example, past trade can be conceptualized as a network of individual entities connected by shared similarity, the flow of goods, to represent their interactions [@Collar2015, pp. 4]. Similar concepts can be applied to a wide range of archaeological data with relational assumptions, such as burial contexts. Burial goods, especially high value goods, can reflect social practices in broader cultural contexts to represent personal wealth or social status from which we can infer social differentiation or complexity [@Gamble2002; @Janes2013]. Burials with the same prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This enables the exploration of the structure of the past social organization through the identification of the relationships among burials.

@@

```{r, burial-map, fig.cap = "Map illustrating the location of burials at Kiwulan (suare size is 4 X 4 meters)", fig.width = image_width_2_col_inch}
knitr::include_graphics(here::here("analysis", "figures", "burial-map.png"))
```

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

## Model specification in a Bayesian framework

```{r ergm-terms-table}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
knitr::kable(ergm_terms,
             caption = "The parameters of exponential random graph models for a undirected network to corresponding with archaeological evidence.")
```

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European percicolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region [@Acabado2017; @Trabert2017]. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.
